---
title: "Practical 4 Answers"
output: html_document
date: "2024-11-11"
---
# Introduction
In many species, testosterone (T) levels are positively correlated with male reproductive success. For example, males with high levels of T may be larger, more aggressive and develop more elaborate secondary sexual characters. As a consequence, they will be more successful in male-male competition, and more attractive to females. However, if there are such obvious benefits to high T levels, why do we still see substantial variation among individuals? One appealing and often hypothesised explanation for the maintenance of variation in T, and in fitness-related traits in general, is that there are not only benefits, but also downsides to high T levels, i.e. there are trade-offs. For example, T may have immunosuppressant effects and increase an individual’s susceptibility to parasites.

Despite a substantial body of work on the role of androgens in mediating a trade-off between reproductive success and health in males, Smyth et al. (2016) set out to investigate its role in female meerkats (Suricata suricatta), a cooperatively breeding species. In this species, dominant females have significantly higher levels of androgens than subordinates. Although dominant females have a dramatically higher reproductive success, do they suffer from higher levels of parasitism?

# 2 Getting Started
```{r}
getwd()
setwd("/Users/carlaleone/Desktop/Exeter/Practical 4")
t.data <- read.table("Dataset.csv", header=TRUE, sep = ",",
                      stringsAsFactors = FALSE)
View(t.data)

#changing colnames
colnames(t.data)[11:16] <- c("strong", "toxo", "oxy", "pseuda", "spiru", "cocci") 
colnames(t.data)
```
## 2.2 Descriptive stats
Getting number of observations per individuals
```{r}
table(t.data$individual)

# also want to know how many individuals have been included how many times:
counts.id <- table(t.data$individual)
table(counts.id)
```

### Excercise 1
Confirm the sample sizes provided in section 2a of the paper (‘Study site and subjects’) using length(), table() and unique(). The latter removes any duplicates from a vector or data frame.
```{r}
length(unique(t.data$group.size))
# 20 groups of different sizes. 
length(unique(t.data$individual))
# 37 individuals
table(unique(t.data$status))
#  D  S 
# 17 38 
# might be repetitions of individuals

dominance<- t.data[!duplicated(t.data$individual), ]
table(dominance$status)
       
```

## 2.3 Parasite abundance and richness
Abundance of each of the parasites is scored on a scale from 0 to 4. Trying to plot the distribution of discrete count data using hist() often doesn’t works so well. For example, let’s visualise the parasite abundance for strong using hist():
```{r}
hist(t.data$strong)
# using a table instead
table.strong <- table(t.data$strong)
barplot(table.strong, xlab="number of eggs", ylab="frequency",
main = "Strongylates",
names.arg=c("0", "1", "2-7", "7-20", ">20"))
```
### Excercise 2
Discuss the pros and cons of the authors’ measure of parasite abundance. And do you think it is appropriate to analyse this with a GLM with Poisson errors?

The bins are not representing equal numbers of eggs. It seems to skew the data a bit. It does help in condensing the really large values, which may reduce the number of outliers. 

Parasite species richness (PSR) obtained by counting parasite taxa.

# 3 Variation in faecal androgen metabolites (FAM)

Rather than measuring the level of T in the blood, the authors measured the amount of androgen metabolites in faecal samples (FAM). Before we start looking at its relationship with parasite presence, diversity and abundance, we should have a closer look at FAM itself. Do we find that dominant females have higher levels of FAM?

## 3.1 cHECK FOR NORMALITY
```{r}
par(mfrow=c(1,2)) 
hist(t.data$fam, main=NULL) 
qqnorm(t.data$fam, main = NULL) 
qqline(t.data$fam)
```
Is not normal, can check with Shapiro test
```{r}
shapiro.test(t.data$fam)
# if significant means not normal
```
## 3.2 Transformations
