---
title: "Tutorial 5 Notes"
output: html_document
date: "2024-11-27"
---

# Intro
Combining raster and vector dta, focus onhow to create or modify raster data using vectors.
Use cropping, masking, rasterization, feature distance grids.
Using same data from last tutorial, with the fires in a PA in Madagascar. There the data were stored as rasters, but that is not a very efficient way. Forest cover was in raster form. but here we generate the same outputs using the vector-raster outputs.

## Load packages
```{r}
library(sf) 
library(tidyverse) 
library(terra) 
library(tidyterra) 
library(colorspace)

getwd()
setwd("/Users/carlaleone/Desktop/Exeter/GIS/Week 10/Tutorial 5")
```

Additional class of spatial data. So far we have been using sf() for vector objects and tera() for raster data. But tera has some limitations and has its own spatial vector format which needs to be used to interact with rasters. 

## Converting to different format
```{r}
# read the rnp boundary
rnp<- st_read("data/ranomafana.gpkg")
#normal simple feature vector format

#convert to the vector format
rnp<- vect(rnp) # now a spat vector object
rnp<- st_as_sf(rnp) # convert back to simple feature
#can also do:
rdp<- vect("data/ranomafana.gpkg")
```

It is very common to convert back and forth between these two vector formats depending on which operation you want to perform (e.g. to sf for geoprocessing and to SpatVector for interactions with raster data). For example, we could transform our national park boundary to projected coordinates (EPSG:32738) using sf then convert back to a SpatVector for interacting with raster data.
```{r}
rnp<- st_as_sf(rnp) %>% st_transform(32738) %>%
vect()
```

We could also convert back to sf for generating our 10 km buffer zone around the national park using st_buffer, although terra also has its own buffer function for buffering SpatVectors
```{r}
#buffer = st_as_sf(rnp) %>% st_buffer(10000) %>% vect()
buffer = buffer(rnp,10000)
```

## Read in forest cover data
```{r}
forest<- rast("data/forest2017.tif")
forest #huge raster
ggplot() +
  geom_spatraster(data=forest) +
  geom_spatvector(data=rnp, fill = "red") +
  geom_spatvector(data = buffer, fill = "transparent", colour = "red")
```
## Clip our raster
raster is very big, so want to crop to our site
Use a pair of terra functions called `crop` and `mask`
```{r}
forest<- crop(forest,buffer) #crop forest to buffer
#now plot is cropped to buffer zone
ggplot() +
  geom_spatraster(data=forest) +
  geom_spatvector(data=rnp, fill = "red") +
  geom_spatvector(data = buffer, fill = "transparent", colour = "red")
```

Next want to mask the buffer zone, because can still see some forest. Mask sets the value of any pictures outside our masking polygon layer to NA. Need to crop before masking, otherwise all the values outside polygon would be NA but it would not have cropped to that location, so would be on way too big of a scale.
```{r}
forest<- mask(forest, buffer)
ggplot() +
  geom_spatraster(data=forest) +
  geom_spatvector(data=rnp, fill = "red") +
  geom_spatvector(data = buffer, fill = "transparent", colour = "red")

#with pipes:
forest<- rast("data/forest2017.tif")%>%
  crop(buffer)%>%
  mas(buffer)
```
## Plotting
Can make it a categorical rather than conitnuous raster by setting levels in the raster
```{r}
levels(forest)<- data.frame(id = 1, name = "forest")

rnp_plot<- ggplot() +
  geom_spatraster(data=forest) +
  geom_spatvector(data=rnp, fill = "transparent") +
  geom_spatvector(data = buffer, fill = "transparent", linetype = "dashed") +
  scale_fill_manual(values = "forestgreen", na.translate = F)

```

## Read in fire data
```{r}
fires<- vect("data/fires.gpkg")
fires #spatvector

rnp_plot +
  geom_spatvector (data = fires, colour = "red")
```
## Rasterize fires
We may want to summarize fire extent, in the last tutorial we made a grid and counted value in the grid cells. But here we will turn our fire point data into a raster layer. This is rasterization:
Use the vector you want to rasterize and a template raster which it will use to count points in the cell. Then need to give it a function for counting the points.
```{r}
fire_grid<- rasterize(fires, forest, fun = length)
#if you look at fires has 30 m by 30 m cells, really small
#need new blank raster, dimensions of buffer, resolution is 1000 x 1000 m
template<- rast(buffer, resolution =c(1000, 1000))
#now rasterize fires onto this template
fire_grid<- rasterize(fires, template, fun = length)



ggplot()+
  geom_spatraster(data = fire_grid)+
  geom_spatvector(data=rnp, fill = "transparent") +
  geom_spatvector(data = buffer, fill = "transparent", linetype = "dashed") +
  scale_fill_continuous_sequential("Heat", na.value = "transparent")

#plot brightness to see how big the fres were. Is a column in the data set.
fire_bright<- rasterize(fires, template,field = 'brightness',fun = sum) 
```

finding number of fires within 1 km of each grid cell
```{r}
#first buffer each cell
fire_buffer<- buffer(fires, 1000)

rnp_plot +geom_spatvector(data=fire_buffer)
#rasterize onto the forest grid

fire_bufgrid<- rasterize(fire_buffer, forest, sum=T)
#dont use the function argument anyomore, because these ar no longer points but polygons

fire_bufgrid[fire_bufgrid==0] <- NA

ggplot()+
  geom_spatraster(data = fire_bufgrid)+
  geom_spatvector(data=rnp, fill = "transparent") +
  geom_spatvector(data = buffer, fill = "transparent", linetype = "dashed") +
  scale_fill_continuous_sequential("Heat", na.value = "transparent")

  

```

Using distance grids: distance of each cell on a raster to the nearest point on a vector feature. measuring distance of forest filled cells to nearest fire:

# distance to fires
```{r}
fire_distance<- distance(forest, fires)
plot(fire_distance) 
# we want to mask this to just the forested pixels in the raster
```

Need to use the mask function:
```{r}
fire_distance<- mask(fire_distance, forest) #sets any pixels in fire distance that dont contain forest to NA. 
# so only forested pixels now

plot(fire_distance) # each forested cell, how far it was from closest fire. 
```
# highlight areas that fall within 1 km of a fire

first duplicate fire distance
```{r}
d<- fire_distance
d[d<= 1000] <- 1
d[d> 1000] <- 2
plot(d)
# now 1s and 2s are the data. We can no make this a categorical raster by setting the levels, which we do by making a data frame

levels(d) <- data.frame(id = 1:2, label = c("< 1km from fire", "> 1km from fire"))
plot(d)
```

plot with ggplot
```{r}
ggplot()+
  geom_spatraster(data = d)+
  geom_spatvector(data=rnp, fill = "transparent") +
  geom_spatvector(data = buffer, fill = "transparent", linetype = "dashed") +
  geom_spatvector(data=fires, size = 0.1) +
  scale_fill_manual (values = c("orange", "forestgreen"), na.translate =FALSE) #categorical so manual

```

Now measuring proportion  of forested area within 1km, inside and outside of national park:
```{r}
# first mask the raster to inside and outside the park:
rnp_forest<- mask (d, rnp) #area just in the park
buf_forest<- mask(d,rnp, inverse = TRUE) # Only pixels outside the park

#now make an at risk layer for each zone, duplicate layers:
rnp_risk <- rnp_forest
rnp_risk[rnp_risk==2]<- NA #ANY OUTSIDE OF 1KM ZONE IS NA

buffer_risk<- buf_forest
buffer_risk[buffer_risk==2]<- NA
#any cell > 1km is NA

#now measure areas:
expanse(buffer_risk, unit = "km")/expanse(buf_forest, unit="km")
expanse(rnp_risk, unit = "km")/expanse(rnp_forest, unit="km")
# result is in percentages
```

