---
title: "w8 tutorial 2"
output: html_document
date: "2024-11-12"
---
# Part 1
Getting started with Pink Seafan Distribution data

Set up
```{r}
getwd()
setwd("/Users/carlaleone/Desktop/Exeter/GIS/Week 8/Tutorial 2")

#load packages
library(sf)
library(tidyverse)
library(colorspace)

#load data
fans <- read_csv("data /pink_sea_fans.csv")
fans #not in spatial format, need to make sf
fans<- st_as_sf(fans, coords = c("Longitude", "Latitude"), crs = 4326)
```

Data collected over long period of time. So limit analysis since year 2000
```{r}
range(fans$Start.date.year) 
# 1960 - 2020

#create subset and remove redundant info in data
fans<- subset(fans, Start.date.year >= 2000, select = "Common.name")

fans

```

Now do everything together using pipes:
```{r}
fans<- read_csv("data /pink_sea_fans.csv") %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>%
  subset(Start.date.year >= 2000, select = "Common.name")
```
# 2 Plot data

first load in uk based map
```{r}
uk<- st_read("data /uk.shp")

ggplot()+
  geom_sf(data=uk)+
  geom_sf(data = fans, colour = "red")
```
Look closer to Falmouth Bay
```{r}
fal.bay<- st_read("data /falmouth_bay.shp")
ggplot ()+
  geom_sf(data=fal.bay)
```
Now crop our records to only retain the fal.bay; use st_crop
```{r}
st_crop(fans, fal.bay)
#error for crs not being same, so transform fans onto fal.bay

fans<- st_transform(fans, crs = 27700)

#name new layer
fb.fans<- st_crop(fans, fal.bay)

#now add this to the plot:
fan.plot<- ggplot ()+
  geom_sf(data=fal.bay) +
  geom_sf(data = fb.fans)

fan.plot
# hotspot south
```

# Part 3 Want to measure area of occupancy of pink sea fans. 
*Area of occupancy* = the number of regularly sized grid cells occupied by species.
First need to create an overlay grid of 1 by 1 km  cells over the study area
```{r}
#First need to create an overlay grid of 1 by 1 km  cells over the study area
grid<- st_make_grid(fal.bay, cellsize = 1000)

#add to plot
fan.plot + geom_sf(data=grid, fill = "transparent")

# default is square, but can make hexagons
grid<- st_make_grid(fal.bay, cellsize = 1000, square = FALSE)
fan.plot + geom_sf(data=grid, fill = "transparent")
```
Remove the grid on land, only want in the sea
```{r}
grid<- st_difference(grid, fal.bay) # fal bay over the grid
grid<- st_crop(grid, fal.bay) #crop so grid doesnt go around the map box

fan.plot+ geom_sf(data=grid, fill = "transparent")
```

Looking at grid, want to turn it into a normal spatial data frame so we can use it for other things later.
```{r}
grid # no attritbutes

grid<- st_as_sf(grid)
grid

# use pipes:
grid<- st_make_grid(fal.bay, cellsize = 1000, square= F) %>%
  st_difference(fal.bay) %>%
  st_crop(fal.bay) %>%
  st_as_sf()
```


##  3.1 How many grids have sea fan observations
using *spatial queries* = subsetting one set of geometries based on their spatial relationship with another. 
In R based on *topological relations* -> take two geometries and test whether a condition is met, describing the relationship. Gives a T or F answer.

Here we use st_intersects()
```{r}
st_intersects(grid, fb.fans, sparse = F)

# want to filter out the grid cells/rows that have at least 1 sighting

st_filter(grid, fb.fans) 
# 45 different features filtered

# create a new grid filtered
grid<- st_filter(grid, fb.fans) 

#plot with new grid
fan.plot+ geom_sf(data=grid, fill = "transparent")

#pipe
grid<- st_make_grid(fal.bay, cellsize = 1000, square= F) %>%
  st_difference(fal.bay) %>%
  st_crop(fal.bay) %>%
  st_as_sf() %>%
  st_filter(fb.fans) 

```

Can use many different spatial relations
```{r}
?st_intersects
```

## 3.2 Remove outliers
removing dodgy observation > 4 km from shore
```{r}
st_filter(fb.fans, fal.bay, .predicate = st_is_within_distance, dist = 4000)
# removed 2 points
fb.fans<- st_filter(fb.fans, fal.bay, .predicate = st_is_within_distance, dist = 4000)

#remake fan plot without dodgy points
fan.plot<- ggplot ()+
  geom_sf(data=fal.bay) +
  geom_sf(data = fb.fans)

fan.plot 

#remake grid and add to plot
grid<- st_make_grid(fal.bay, cellsize = 1000, square= F) %>%
  st_difference(fal.bay) %>%
  st_crop(fal.bay) %>%
  st_as_sf() %>%
  st_filter(fb.fans) 

fan.plot+ geom_sf(data=grid, fill = "transparent")
```

Use the st_area function to calculate the area of occupancy
```{r}
st_area(grid)

```
Some cells have more observations than others.
Want to count number of observations per cell:
```{r}
st_intersects(grid, fb.fans)
# output = list of IDs of each point that intersects each grid cell.

# need to count id for each element of the list:
grid$number_of_fans<- lengths(st_intersects(grid, fb.fans))
grid
```
Plot the grid, colour based on number of sightings:
```{r}
ggplot()+
  geom_sf(data = fal.bay) +
  geom_sf(data =grid, mapping = aes(fill = number_of_fans)) +
  scale_fill_continuous_sequential(palette = "RdPu")

#  hcl_palettes(plot = T) to find the colour palettes available
```

# Part 4 MPAs

read mpa boundaries
```{r}
mcz<- st_read("data /MCZs.gpkg")
sac<- st_read("data /SACs.gpkg")

# add to fan plot
fan.plot +
  geom_sf(data =sac, colour ="blue") +
  geom_sf(data = mcz, colour = "green")
```
Count number of seafans in each mpa;
```{r}
st_filter(fb.fans, mcz)
#extracted no features

mcz
# only line features. none intersect the observations. 
# Convert MPA boundaries to polygons
mcz<- st_cast(mcz, "MULTIPOLYGON")
sac<- st_cast(sac, "MULTIPOLYGON")

# Use pipes:
mcz<- st_read("data /MCZs.gpkg") %>%
  st_cast( "MULTIPOLYGON")

sac<- st_read("data /SACs.gpkg") %>%
  st_cast( "MULTIPOLYGON")

fan.plot<- fan.plot +
  geom_sf(data =sac, fill ="blue", alpha = 0.2) +
  geom_sf(data = mcz, fill = "green", alpha = 0.2)

```

Now filter: fans that occur in MPAs:
```{r}
st_filter(fb.fans, mcz) #220
st_filter(fb.fans, sac) #73

(220+73)/nrow(fb.fans)
#62.8% of sightings occur in an MPA
```

# Part 5 Limitations:
Sightings are biased to areas with frequent visits. IE dive sites in the south.
Could look at how well MPAs are protecting sea fan habitats.

Need to know about sea fan habitat preferences:
```{r}
habitats<- st_read("data /uk_seamap.gpkg")
habitats 
# has a lot of info . We only care about JNCC name:

habitats<- st_read("data /uk_seamap.gpkg") %>%
  subset(select = "JNCCName") %>%
  rename("habitat" =  JNCCName) %>%
  drop_na(habitat)

habitats
```

Plot the data:
```{r}
fan.plot+
  geom_sf(data=habitats, mapping = aes(fill = habitat))
```
Many polygons for each habitat type. Have multiple polygons per habitat type, so would be better to dissolve these so we have a single feature for each habitat type.
Use st_union:
```{r}
group_by(habitats, habitat) %>%
  summarise(do_union =T)
#13 features, 1 polygon per habitat type

#with pipes
habitats<- st_read("data /uk_seamap.gpkg") %>%
  subset(select = "JNCCName") %>%
  rename("habitat" =  JNCCName) %>%
  drop_na(habitat) %>%
  group_by( habitat) %>%
  summarise(do_union =T)

fan.plot+
  geom_sf(data=habitats, mapping = aes(fill = habitat)) +
  geom_sf(data = fb.fans)
```

Extract habitat in which each seafan is located. Use *spatial join* = joins attributes of one layer to another based on geometric relationship between the two.
st_join by default uses intersects for joining.
```{r}
fb.fans<- st_join(fb.fans, habitats)
#now each observation has a habitat added to it.

#what is the species habitat preference?
table(fb.fans$habitat)
#counts how many times observations appear for each habitat. 
#Could work out proportions
table(fb.fans$habitat)/nrow(fb.fans)
# 54.5% from  Moderate energy circalittoral rock

# bar plot
ggplot() +
geom_bar(data =fb.fans, mapping = aes(x=habitat)) +
  coord_flip()
```
Possible that this habitat just makes up majority of fal bay area. Check area of each habitat:
```{r}
habitats$proportion<- st_area(habitats) / sum(st_area(habitats))
habitats
# only makes up 5% of seabed area but accounts for 54% of sea fan observations
# suggests strong preference. but ideally would do formal species distribution model
```
# 5.1 What proportion of this habitat is in MPAs?
```{r}
rock<- subset(habitats, habitat == "Moderate energy circalittoral rock")

fan.plot +
  geom_sf(data = rock, fill ="khaki4", alpha = .5)

```
Rocky habitats seem to nicely overlap with observations. 
Now lets see how much is in the MPAs:

```{r}
# cut out bits of rock that fall into the MPAs. Use st_intersection (check document)
mcz_rock<- st_intersection(rock, mcz)
sac_rock<- st_intersection(rock, sac)

fan.plot +
  geom_sf(data = mcz_rock, fill ="khaki4", alpha = .5)

st_area(mcz_rock)
# 837406.6 [m^2]

st_area(sac_rock)
# 1483984 [m^2]

# total proportion of rock in the MPAs
st_area(mcz_rock)/st_area(rock)
# 0.04480402 [1]

st_area(sac_rock)/st_area(rock)
# 0.07939804 [1]

```

In total around 12% of the rock habitat is in the MPAs.

