---
title: "Tutorial 3 Notes"
output: html_document
date: "2024-11-20"
---

Packages:
```{r}
library(sf)
library(tidyverse)
library(terra)
install.packages("tidyterra")
library(tidyterra)
library(colorspace)

getwd()
setwd("/Users/carlaleone/Desktop/Exeter/GIS/Week 9/Tutorial 3")
```

Importing raster data
```{r}
dem <- rast("data/dem.tif")
dem

#other functions:
ncell(dem) # number of cells
res(dem) # resolution of the raster
ext(dem)#spatial extent
names(dem)
```
spat raster = spatial raster. tells us we have ... rows and columns, how many layers, the resolution, spatial extent

# Plot the elevation data
```{r}
plot(dem)
```
plot(dem) is optimized for raster data, but still lacks versatility of other dedicated plotting packages like ggplot

```{r}
ggplot()+
  geom_spatraster(data = dem) #dont need to specify fill aesthetics
```
ggplot plots missing values in grey as default. But we want to change the background colour and the gradient. so need to apply a custom fill scale

```{r}
hcl_palettes(plot=TRUE) #available palettes
ggplot()+
  geom_spatraster(data = dem) +
  scale_fill_continuous_sequential("Terrain 2", na.value = "transparent", rev= FALSE) # colour palette by default is usally reversed, so need to make sure that higher elevation corresponds to less green
```

For elevation we can use other palettes from hypsometric tints in tidyterra
```{r}
dem_plot<- ggplot()+
  geom_spatraster(data = dem) +
  scale_fill_hypso_c("wiki-2.0_hypso") #na is automatically transparent
```


# Plot leopard data
leopards also have a column called sample which specifies if sample was from cam trap or observation
```{r}
leopards<- st_read("data/leopards.gpkg")
ladakh<- st_read("data/ladakh.gpkg")

# plot new
dem_plot +
  geom_sf(data=leopards, mapping = aes(colour = sample, shape = sample)) +
  geom_sf(data = ladakh, fill = "transparent")
```
#Summarize elevation
Summary stats of the new raster layer. use global statistsics
```{r}
global(dem, fun = "mean", na.rm = TRUE) 
#MEAN ELEVALTION = 4846.711	

global(dem, fun = "min", na.rm = TRUE) 

minmax(dem) # gives min and max together
#range of elevation is 2575 m  to 7181 m
```

Focal statistics, calculate new value in a cell based on the values of the cells around it. Returns a new raster layer.
The focal stats we need are already coded, but can check the reading. Common use is terrain analysis:
ruggedness, slope angle, aspect ... all habitat variables used in the study

# Terrain analysis
```{r}
slope<- terrain(dem, v = "slope")
plot(slope)

ggplot()+
  geom_spatraster(data = slope) +
  scale_fill_continuous_sequential("Heat", na.val = "transparent")
```

Next calculate terrain rugedness:
```{r}
TRI<- terrain(dem, v = "TRI")

ggplot()+
  geom_spatraster(data = TRI) +
  scale_fill_continuous_sequential("Turku", na.val = "transparent")
```
Aspect = direction that slopes face
```{r}
aspect<- terrain (dem, v = "aspect")
ggplot()+
  geom_spatraster(data = aspect) +
  scale_fill_continuous_sequential("Sunset", na.val = "transparent")
```
Aspect is hard to plot becuase is circular, measured in degrees 0-360. So we may want to split between N,E,S,W
Use raster classification

#Calssify elevation rasters
involves grouping cells into defined classes and then relabelling them as something else. in Terra done with the classify function
```{r}
classify(dem, rcl = seq(2000, 8000, 1000)) %>% plot #cells between 2000 and 8000 m and classify every 1000 m 
```
Need to build a relcassification table, to give the ranges for each class and what we want them relabled as:
```{r}
# Build a numeric object how you want your table to look, and then wrap in a matrix
rcl<- matrix(
c(0, 45, 1, #north
  45,135, 2, #east
  135, 225, 3, #south 
  225, 315, 4, #west
  315, 360, 1),
ncol = 3, byrow = T) #back to north

aspect<- classify(aspect, rcl = rcl)
plot(aspect)

ggplot()+
  geom_spatraster(data = aspect) +
  scale_fill_continuous_sequential("Sunset", na.val = "transparent")
# now the classes dont have names, but we can give names to the classes through a raster attribute table
```
Making an attribute table for the aspect raster. Tner first column needs to be called ID and needs to conatin the vaue in the raster. Our values are 1-4. Then we can add any other columns for how those values map to categorical variables:
```{r}
rat<- data.frame(id = 1:4, aspect = c("N", "E", "S", "W"))
levels(aspect) <- rat
#levels in aspect are the values in our attribute table

#now our scale is discrete and not continuous, so need to change the scale_fill
ggplot()+
  geom_spatraster(data = aspect) +
  scale_fill_discrete_sequential("Sunset", na.translate = F) #remove na
```
# Suitable Habitat
Authors of the paper reported that suitable habitat for snow leopards was between 2800 and 4600 m. So we can classify all the habitat that is within that elevation range:

Can test whether each cell is lower than 4600  and higher than 2800:
```{r}
plot(dem < 4600)
plot(dem > 2800)

#combine statements
plot(dem > 2800 & dem < 4600)
suitable<- dem > 2800 & dem < 4600
```

Subset the suitable parts and find the area:
We can do this by just keeping the cells with the values we want and making everything else in the raster to NA
```{r}
suitable[suitable == 0] <- NA
# Any cells in suitable that are 0, should be NA
plot(suitable)

#now measure the total area 
expanse(suitable, unit = "km")
# 20,683 km ^2 of suitable habitat in Ladakh

#calculate proportion of suitable area
expanse (suitable)/expanse(dem)
# around 32.8% of habitat in ladakh is suitable

`````````
# Export new raster

```{r}
writeRaster(suitable, "suitable_habitat.tif")
```

